# Results

```{r, message= FALSE, warning=FALSE}
# Importing library
library(tidyverse)  
library(readr)      
library(ggplot2)    
library(dplyr)
library(gridExtra)

tree_data <- read_csv("../2015_Street_Tree_Census_-_Tree_Data_20241120.csv", n_max = 266134)
tree_data <- na.omit(tree_data)
```
## The Urban Canopy: NYC’s Tree Diversity
Lets first explore the overall composition of NYC's tree population.

```{r, fig.width=15, fig.height=10}

# Count the number of trees for each species and select the top 15
top_species <- tree_data %>%
  count(spc_common, sort = TRUE) %>%
  slice_max(order_by = n, n = 15)  

# Count the number of trees for each species and select the least 15
least_species <- tree_data %>%
  count(spc_common, sort = TRUE) %>%
  slice_min(order_by = n, n = 15) 

# Create the bar plot for Top 15 Most Common Tree Species
top_species_plot <- ggplot(top_species, aes(x = reorder(spc_common, n), y = n)) +
  geom_bar(stat = "identity", fill = "#4C72B0") + 
  coord_flip() + 
  labs(
    title = "Top 15 Most Common Tree Species in NYC", 
    x = "Tree Species", 
    y = "Number of Trees"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),  
    axis.text.y = element_text(face = "bold", size = 12),  
    axis.text.x = element_text(face = "bold", size = 12),  
    axis.title = element_text(face = "bold", size = 14))

# Create the bar plot for 15 Least Seen Tree Species
least_species_plot <- ggplot(least_species, aes(x = reorder(spc_common, n), y = n)) +
  geom_bar(stat = "identity", fill = "#D55E00") +  
  coord_flip() + 
  labs(
    title = "15 Least Common Tree Species in NYC", 
    x = "Tree Species", 
    y = "Number of Trees"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),  
    axis.text.y = element_text(face = "bold", size = 12),  
    axis.text.x = element_text(face = "bold", size = 12),  
    axis.title = element_text(face = "bold", size = 14)  
  )

# Arrange the plots side-by-side using gridExtra
grid.arrange(
  top_species_plot, 
  least_species_plot, 
  ncol = 2  
)

```
The bar plot highlights that the **London planetree** is the most common species in NYC, with over 30,000 trees, thanks to its adaptability to urban conditions such as tolerance to pollution, drought, and varying soil pH. While it contributes to NYC’s fall aesthetics with its dense canopies and abundant leaf shedding, its foliage lacks the vibrant colors of species like **red maple** and **sweetgum**, which define the iconic autumn palette. Other species like **honeylocust**, **Callery pear**, and **pin oak** also play significant roles in NYC’s tree population. However, the heavy reliance on the London planetree raises concerns about monoculture vulnerabilities, as pests or diseases could disproportionately affect it. Insights from this analysis suggest that diversifying species further could enhance ecological resilience and reduce maintenance challenges, such as managing the significant cleanup required for London planetree debris.

The second bar plot shows that NYC’s **least common species**, such as **red pine**, **red horse chestnut**, and **Douglas-fir**, have fewer than 30 trees. Their rarity reflects challenges in urban adaptation, including specific environmental needs and sensitivity to pollution. Trees like **Osage-orange** and **Himalayan cedar** highlight underutilized biodiversity, offering potential for conservation efforts. Rare species may require tailored strategies for survival in urban environments, such as improved soil preparation, reduced competition, or protection from pests. Incorporating a wider variety of species, including these rarer types, could reduce risks associated with monoculture and promote a more balanced and resilient urban forest.

## Surviving the City: Daily Struggles of NYC Trees

We will see the prevalence of various tree problems in NYC, categorized into trunk, branch, and root issues using the faceted bar chart.
```{r, fig.width=15, fig.height=10}
problem_columns <- c("trunk_wire", "trnk_light", "trnk_other", 
                     "brch_light", "brch_shoe", "brch_other",
                     "root_stone", "root_grate", "root_other")

# Prepare data
problem_data <- tree_data %>%
  select(all_of(problem_columns)) %>%
  pivot_longer(cols = everything(), names_to = "problem_type", values_to = "presence") %>%
  filter(!is.na(presence)) %>%
  group_by(problem_type, presence) %>%
  summarise(count = n(), .groups = "drop")

# Create the faceted bar chart
ggplot(problem_data, aes(x = presence, y = count, fill = presence)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("Yes" = "#FF6347", "No" = "#4682B4")) +
  facet_wrap(~ problem_type, scales = "free_y") +
  labs(
    title = "Presence of Tree Problems by Type",
    x = "Problem Presence (Yes/No)",
    y = "Number of Trees"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold", size = 14)
  )


```

**Branch Problems**

- **Branch problems caused by lights** (`brch_light`) have the highest occurrence among branch-related issues but remain rare compared to "No" responses.
- Problems like **branch problems caused by shoes** (`brch_shoe`) and **other branch problems** (`brch_other`) are negligible.

**Root Problems**

- **Root problems caused by stones** (`root_stone`) are the most significant among all categories, with a notable number of "Yes" responses. This highlights the impact of paved areas on root health.
- **Root problems caused by grates** (`root_grate`) and **other root problems** (`root_other`) are nearly nonexistent.

**Trunk Problems**

- Trunk-related issues, including **problems caused by wires** (`trunk_wire`) and **lights** (`trnk_light`), are minimal across all trees.
- **Other trunk problems** (`trnk_other`) are extremely rare and show almost no significant impact.

The graph highlights **root problems caused by stones** as the primary concern, underscoring the need for improved urban planning around tree beds. Branch and trunk problems are negligible, indicating effective maintenance or fewer external stressors.


## Rooting for Good Health: Tree Conditions in NYC Neighborhoods
The stacked diverging bar chart provides a visual summary of tree health distribution.
```{r}
# Prepare data
tree_data_filtered <- tree_data %>%
  filter(!is.na(health), !is.na(borough)) %>%  
  group_by(borough, health) %>%
  summarise(count = n(), .groups = "drop") %>%
  mutate(
    health = factor(health, levels = c("Poor", "Fair", "Good"))  
  )

# Plot data
ggplot(tree_data_filtered, aes(x = count, y = borough, fill = health)) +
  geom_bar(stat = "identity", position = "stack") +  
  scale_fill_manual(
    values = c("Poor" = "#F44336", "Fair" = "#FFC107", "Good" = "#4CAF50"),
    name = "Tree Health"
  ) +
  labs(
    title = "Tree Health Distribution by Borough",
    x = "Number of Trees",
    y = "Borough"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.title = element_text(face = "bold", size = 12))

```

**Overall Tree Health**

- The majority of trees in all boroughs are classified as being in **Good health** (green segment), indicating a generally healthy urban forest in NYC.

**Borough Comparison**

- **Queens** has the highest number of trees overall, with a significant portion in good health, but it also has a noticeable proportion of trees in **Fair health** (yellow segment).
- **Staten Island** and **Bronx** have relatively fewer trees compared to Brooklyn and Queens, but a large proportion of their trees are in **Good health**.

**Poor Health Distribution**

   - Trees in **Poor health** (red segment) represent the smallest category across all boroughs.
   - **Manhattan** has the highest proportion of trees in poor health relative to its total number of trees, suggesting potential urban stress factors like pollution or limited green spaces.

**Brooklyn and Staten Island**

   - Brooklyn has the second-highest total number of trees, following Queens, and most are in good health.
   - Staten Island, despite having the smallest tree count, shows a relatively balanced distribution among health categories compared to other boroughs.

**Potential for Action**

   - Boroughs like **Manhattan** and **Queens** may require targeted intervention to improve the health of trees classified as **Fair** or **Poor**, given their larger populations and stress factors.
   
## Trunk Titans: Small vs. Mighty Trees
The Cleveland dot plot shows the average tree diameter (DBH) for various tree species, highlighting differences in growth patterns and size among species in NYC.
```{r, fig.width=17, fig.height=12}
library(dplyr)
library(ggplot2)

# Aggregate mean diameter by species
species_avg <- tree_data %>%
  group_by(spc_common) %>%
  summarise(mean_dbh = mean(tree_dbh, na.rm = TRUE)) %>%
  arrange(desc(mean_dbh)) %>%
  slice(1:50) 

# Create the plot
ggplot(species_avg, aes(x = mean_dbh, y = reorder(spc_common, mean_dbh))) +
  geom_point(color = "blue", size = 3) +
  labs(
    title = "Average Tree Diameter by Species",
    x = "Average Diameter (DBH)",
    y = "Tree Species"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    axis.text.y = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.title.x = element_text(face = "bold", size = 12),
    axis.title.y = element_text(face = "bold", size = 12)
  )


```

**Top Species with Largest Average Diameters**

   - Species like **London planetree**, **silver maple**, and **weeping willow** stand out with the largest average tree diameters, exceeding 16 DBH (Diameter at Breast Height). These species are known for their fast growth and adaptability, often resulting in larger tree sizes.

**Species with Smallest Average Diameters**

   - At the lower end of the spectrum, species such as **Ohio buckeye**, **ginkgo**, and **American linden** exhibit smaller average diameters (around 8 DBH). These species may be slower-growing or less suitable for NYC's urban conditions.

**Insights for Urban Planning**

   - Large-diameter species like the **London planetree** may require more space and resources but provide substantial canopy coverage. In contrast, smaller-diameter species may be suitable for tighter urban spaces.

**Species-Specific Growth Patterns**

   - The differences in average diameter could reflect species-specific growth characteristics, planting preferences, or environmental adaptability in NYC.
   
## Paving the Way: Proportion of Sidewalks Affected by Trees
The pie chart illustrates the proportion of trees in NYC with adjacent sidewalk damage.
```{r}
# Calculate proportions
sidewalk_proportion <- tree_data %>%
  count(sidewalk) %>%
  mutate(prop = n / sum(n))

# Enhanced pie chart
ggplot(sidewalk_proportion, aes(x = "", y = prop, fill = sidewalk)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  geom_text(
    aes(label = scales::percent(prop, accuracy = 0.1)),
    position = position_stack(vjust = 0.5),
    color = "white",
    size = 5
  ) +
  labs(
    title = "Proportion of Trees with Sidewalk Damage",
    fill = "Sidewalk Condition"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    legend.position = "right",
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 10)
  ) +
  scale_fill_manual(values = c("Damage" = "#F44336", "NoDamage" = "#4CAF50"))

```

**Majority with No Sidewalk Damage**:

   - A significant proportion (71.5%) of trees are adjacent to sidewalks that are in good condition, without visible damage.

**Minority with Sidewalk Damage**:

   - Approximately 28.5% of trees are associated with sidewalks showing signs of damage.

**Urban Impact Insight**:

   - This suggests that while the majority of sidewalks are resilient to tree-related damage, a substantial portion still suffers from issues, potentially due to tree roots disrupting pavement structures.

**Actionable Consideration**:

   - Efforts could focus on mitigating damage through better tree placement or root management strategies to reduce future sidewalk damage while preserving urban greenery.

## Pathways to Tree Health: Exploring Curb Placement, Sidewalk Conditions, and Their Impact
The alluvial plot visualizes the relationship between curb location, sidewalk damage, and tree health.
```{r, fig.width=15, fig.height=8}
library(ggalluvial)
# Prepare data for alluvial plot
alluvial_data <- tree_data %>%
  group_by(curb_loc, sidewalk, health) %>%
  summarise(count = n(), .groups = "drop")

# Alluvial plot with improved axis labels
ggplot(alluvial_data, aes(axis1 = curb_loc, axis2 = sidewalk, axis3 = health, y = count)) +
  geom_alluvium(aes(fill = health)) +
  geom_stratum() +
  geom_text(stat = "stratum", aes(label = after_stat(stratum)), size = 4) + 
  labs(
    title = "Flow of Trees from Curb Location to Sidewalk Damage to Health",
    x = "Tree Characteristics",
    y = "Number of Trees"
  ) +
  scale_fill_manual(values = c("Good" = "#4CAF50", "Fair" = "#FFC107", "Poor" = "#F44336")) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(hjust = 1, size = 12), 
    axis.text.y = element_text(size = 12), 
    axis.title = element_text(size = 12, face = "bold"), 
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))

```
**Curb Location**

   - Most trees are located **OnCurb**, indicating that urban tree planting often aligns with curb locations.
   - Trees **OffsetFromCurb** represent a smaller proportion, likely in areas with wider sidewalks or designated tree beds.

**Sidewalk Damage**

   - Trees **OnCurb** are more associated with **Sidewalk Damage**, likely due to root growth interfering with pavement structures.
   - Trees **OffsetFromCurb** show a stronger association with **NoDamage**, suggesting reduced impact on sidewalks.

**Tree Health**

   - The majority of trees are categorized as **Good**, irrespective of their curb location or sidewalk condition.
   - Trees in **Poor** health are more prevalent where **Sidewalk Damage** is present, indicating that environmental stressors may contribute to declining health.

**Relationships**

   - A clear flow is visible where trees **OnCurb** with **Sidewalk Damage** tend to have more **Fair** or **Poor** health. 
   - Trees **OffsetFromCurb** and with **NoDamage** predominantly exhibit **Good** health.

The plot highlights the interplay between tree placement, sidewalk integrity, and tree health. Proper placement (e.g., OffsetFromCurb) could mitigate sidewalk damage and promote better tree health.

```{r}
library(ggplot2)

ggplot(tree_data, aes(x = sidewalk, y = tree_dbh, fill = sidewalk)) +
  geom_violin() +
  labs(
    title = "Tree Diameter and Sidewalk Damage",
    x = "Sidewalk Condition",
    y = "Tree Diameter (DBH)"
  ) +
  scale_fill_manual(values = c("Damage" = "#F44336", "NoDamage" = "#4CAF50")) +
  coord_cartesian(ylim = c(0, 100)) +  # Adjust the y-axis limits to zoom in
  theme_minimal()

```
### **Bubble chart**

The bubble chart offers several insights into the distribution of tree species and their health status across different boroughs in New York City.

```{r}
library(ggplot2)
library(dplyr)

# Calculate the Top 5 Most Common Species
top_species <- tree_data %>%
  count(spc_common, sort = TRUE) %>%
  top_n(5, wt = n) %>%
  pull(spc_common) # Extract species names as a vector

# Prepare Data for Bubble Chart
bubble_data <- tree_data %>%
  group_by(borough, spc_common, health) %>%
  summarize(count = n(), .groups = "drop") %>%
  filter(spc_common %in% top_species) # Focus on top 5 species

# Bubble Chart
ggplot(bubble_data, aes(x = borough, y = spc_common, size = count, color = health)) +
  geom_point(alpha = 0.7) +
  scale_size_area(max_size = 15) +
  scale_color_manual(values = c(
    "Good" = "#27ae60", 
    "Fair" = "#f1c40f", 
    "Poor" = "#e74c3c"
  )) +
  labs(
    title = "Tree Species vs. Borough by Health",
    x = "Borough",
    y = "Species",
    size = "Tree Count",
    color = "Health Status"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )


```

A few key patterns emerge from this visualization:

1.  **Dominance of London Planetree**:
    -   The London planetree stands out as the most commonly planted species across all boroughs, evidenced by its larger bubble sizes. This aligns with its popularity as a resilient urban tree, often chosen for its ability to thrive in diverse environmental conditions.
2.  **Health Trends Across Species**:
    -   Most trees across all species and boroughs are in good health (represented by the predominance of green bubbles). However, there are smaller proportions of trees in fair (yellow) or poor (red) health, indicating areas that may require targeted maintenance efforts.
3.  **Borough-Specific Characteristics**:
    -   Manhattan has a noticeably high count of London planetrees in good health, making it a significant contributor to the city’s urban forest. Similarly, Brooklyn and Queens also show a strong presence of Callery pear and Norway maple, with most of these trees in good condition.
    -   Staten Island, on the other hand, has fewer trees overall, as reflected in the smaller bubble sizes across all species. This could suggest either less dense tree planting efforts or smaller urban areas requiring greenery.
4.  **Species in Poor Health**:
    -   A few instances of trees in poor health, particularly Norway maple in Queens, stand out in the visualization. These trees may require closer attention to improve their condition and ensure they continue contributing positively to the urban ecosystem.
5.  **Species Diversity**:
    -   While the London planetree dominates, other species such as Callery pear and honeylocust also have significant representation. This diversity is particularly evident in boroughs like Queens, which seem to have a more balanced mix of species compared to other boroughs.



```{r, fig.width=17, fig.height=5}
# Prepare Data for Scatter Plot
scatter_data <- tree_data %>%
  filter(tree_dbh > 0) %>%
  pivot_longer(
    cols = starts_with("root") | starts_with("trunk") | starts_with("brch"), 
    names_to = "problem_type", 
    values_to = "has_problem"
  ) %>%
  filter(!is.na(has_problem) & has_problem == "Yes")

ggplot(scatter_data, aes(x = tree_dbh, y = health, color = borough)) +
  geom_jitter(alpha = 0.3, size = 1.5) +
  facet_wrap(~problem_type, ncol = 4, nrow = 2, scales = "free_y") + # 4 columns and 2 rows
  scale_x_continuous(
    trans = scales::log_trans(base = 2),
    breaks = c(1, 2, 4, 8, 16, 32, 64), # Log2 breaks
    labels = c("1", "2", "4", "8", "16", "32", "64")
  ) +
  scale_color_brewer(palette = "Set2") +
  labs(
    title = "Tree Diameter vs. Health by Problem Type",
    x = "Tree Diameter (DBH) (Log2 Scale)",
    y = "Health Status",
    color = "Borough"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    legend.position = "top",
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
### **Observation and Story: Root-Related Challenges for NYC's Trees**

#### **Observation**
- Trees with root-related problems (`root_grate`, `root_other`, `root_stone`) display a significant decline in health as their diameter (DBH) increases.
- Trees larger than 16 inches (DBH) show a higher proportion of "Fair" and "Poor" health statuses compared to smaller trees.
- This pattern is most prominent in boroughs like **Brooklyn** and **Staten Island**, which have a greater number of larger trees and infrastructure-related stressors.
- Root-related problems, such as interference from grates, stones, and soil compaction, are particularly impactful for larger trees, likely due to their greater resource requirements and restricted growing conditions in urban environments.

#### **Story**
- **Larger Trees Are Under Threat**:
  - Larger trees, especially those exceeding 16 inches in diameter, are critical to NYC's urban canopy. They provide a range of ecological benefits, including carbon sequestration, air purification, cooling through shade, and improved biodiversity. However, the data shows they are increasingly vulnerable to health issues, particularly root-related stress.
  - Root issues, including limited space for root expansion, interference from grates, and urban soil compaction, disrupt nutrient absorption and overall stability, contributing to a decline in health.

- **Brooklyn and Staten Island as Vulnerable Zones**:
  - These boroughs host a broader distribution of larger trees, making them key areas where root-related challenges are more prevalent. Larger trees in these boroughs are often subject to urban stressors due to aging infrastructure and limited urban planning geared toward tree preservation.
  - The prevalence of root issues in these boroughs highlights the need for borough-specific interventions, such as redesigning urban layouts and expanding tree planting zones.

```{r fig.width=9, fig.height=6}
library(ggplot2)
library(scales)

# Custom labels for log base 2
log2_labels <- function(x) {
  paste0("2^", log2(x))
}

# Prepare Data for Stewardship Analysis
steward_health_data <- tree_data %>%
  filter(!is.na(steward)) %>%
  group_by(steward, health) %>%
  summarize(count = n(), .groups = "drop")


# Stacked Bar Chart with Log Base 2 Transformation and Labels
ggplot(steward_health_data, aes(x = steward, y = count, fill = health)) +
  geom_bar(stat = "identity", color = "black", position = "stack") +
  scale_y_continuous(
    trans = scales::log_trans(base = 2), # Logarithmic transformation with base 2
    breaks = scales::trans_breaks("log2", function(x) 2^x), # Breaks at powers of 2
    labels = log2_labels # Custom labels in the form of 2^n
  ) +
  scale_fill_manual(values = c(
    "Good" = "#27ae60", 
    "Fair" = "#f1c40f", 
    "Poor" = "#e74c3c"
  )) +
  labs(
    title = "Tree Health by Stewardship Level (Log Base 2 Scale)",
    x = "Stewardship Level",
    y = "Number of Trees (Log Base 2 Labels)",
    fill = "Health Status"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12), # Larger font size for y-axis labels
    legend.position = "top"
  )



```

### Observation:

- The **stewardship level** significantly influences the health of trees, as evidenced by the data. The graph, plotted on a **log base 2 scale**, highlights that trees without any recorded stewardship ("None") constitute the largest group. This category also exhibits the widest variability in health status, with a substantial proportion of trees in "Good" health, followed by "Fair" and "Poor" health. This could indicate that while trees without stewardship are at higher risk of deteriorating health, a significant number still maintain good health, potentially due to favorable environmental conditions or maintenance by other means.

- Trees under stewardship levels of "1or2" and "3or4" show similar patterns, with a notable dominance of trees in "Good" health. This suggests that even minimal stewardship efforts play a crucial role in maintaining tree health. However, the diminishing trend in tree count across higher stewardship levels indicates fewer trees fall under intensive care or monitoring ("4orMore"), which might imply resource limitations or a focused approach on only select trees.

- The **log scale representation** emphasizes the exponential difference in tree counts across stewardship levels, making it evident that the majority of trees lack any formal stewardship. This highlights an opportunity for urban forestry programs to engage more stewards or allocate resources to currently neglected trees.

- Overall, the graph suggests that stewardship has a positive impact on tree health, but the lack of stewardship for a significant proportion of trees is an area of concern. This could warrant further investigation or policy recommendations to enhance stewardship participation and ensure a healthier urban forest.

```{r}
# Calculate Shannon Diversity Index for Each Borough

library(vegan)
diversity_data <- tree_data %>%
  group_by(borough, spc_common) %>%
  summarize(count = n(), .groups = "drop") %>%
  group_by(borough) %>%
  summarize(diversity_index = diversity(count))

# Bar Chart
ggplot(diversity_data, aes(x = reorder(borough, -diversity_index), y = diversity_index, fill = borough)) +
  geom_col(color = "black", width = 0.7) +
  scale_fill_brewer(palette = "Set3") +
  labs(
    title = "Tree Diversity Index by Borough",
    x = "Borough",
    y = "Diversity Index",
    fill = "Borough"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    legend.position = "none"
  )

```
### Observation:

The bar chart illustrates the **Shannon Diversity Index** for tree species across different boroughs in New York City. The diversity index, which quantifies the variety and distribution of species, reveals some subtle but meaningful patterns:

1. **Queens** emerges with the highest diversity index, indicating that this borough has the most balanced mix of tree species. This could be due to its larger green spaces or more distributed urban forestry initiatives.

2. **Bronx** and **Brooklyn** follow closely, showing comparable diversity indices. These boroughs likely benefit from active reforestation programs or ecological conditions that support a variety of species.

3. **Staten Island** and **Manhattan** have slightly lower diversity indices. Staten Island's lower population density might explain fewer diverse tree planting efforts. On the other hand, Manhattan's highly urbanized landscape and limited green space could restrict opportunities for introducing diverse tree species.

### Interpretation:

This analysis suggests that **urban planning, tree planting policies, and available green space** play significant roles in influencing tree diversity across boroughs. While the differences are not drastic, the diversity index highlights the relative ecological richness of Queens, Bronx, and Brooklyn compared to the more urbanized Manhattan. These findings could guide future biodiversity-enhancing efforts in areas with lower diversity.
